Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework

### 2023-05-30 manuscript/manuscript.Rmd
minor updates.

This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology, which tends to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords. The Triple P framework aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations. To accomplish these two objectives, we advocate the following five measures as integral components of the Triple P framework: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) normalizing the ethological and ethnographic data collection in experimental projects; 3) encouraging the involvements of avocational as well as novice participants; 4) boosting the collaboration across labs on a global scale; 5) building an open-access repository for data reuse. It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020]. Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to clarify its meaning and demonstrate the necessity and potentials of this framework.

@@

As implied in its name, the implementation of Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality, one of daunting challenges faced in archaeological inference [@hiscock2004; @lake2014; @premoEquifinalityExplanationThoughts2010]. Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends.

>>>>

This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology, which tends to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords. Built upon early intellectual principles and practices in behavioral archaeology [@schiffer2010], the Triple P framework aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations. To accomplish these two objectives, we advocate the following five measures as integral components of the Triple P framework: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) normalizing the ethological and ethnographic data collection in experimental projects; 3) encouraging the involvements of avocational as well as novice participants; 4) boosting the collaboration across labs on a global scale; 5) building an open-access repository for data reuse. It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020]. Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to clarify its meaning and demonstrate the necessity and potentials of this framework.

@@

As implied in its name, the implementation of Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@hiscock2004; @premoEquifinalityExplanationThoughts2010], two daunting challenges in archaeological inference that partially contributed to the discipline-wide paradigm shift in the 1980s [@lake2014: 264-265]. Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends. While we cannot fully solve these two problems and accurately reconstruct the past behavioral processes and intentionalities simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better understand the range of possible combinations of variation at these three levels and thereby evaluate the probability of certain behavioral mechanisms behind a given archaeological assemblage.
